Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-08-01 paint-contour-tree.tex
Small changes, section 5.2 not finished

@@

\begin{lemma}
\label{lem:pathBounds}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Then, 
 \begin{compactenum}
  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
  \item for any light path $p\in P(T)$, $\sum_{v\in P} |H_v| = O(|p| \log |H_{r_p}|)$.
 \end{compactenum}
\end{lemma}
\begin{proof}
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
 As such, 
 \[
 \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
 \]
 If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
 $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
\end{proof}

@@

 Let $H_i$ be the heap at $r_{q_i}$ and let $H_0$ be the heap at $r_p$.  Then $\sum_{i=1}^k |H_i| \leq |H_0|+|p|$.
 %Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
\end{lemma}
\begin{proof}
 %Let $c_1, \dots, c_k$ be the connections of the paths $q_1, \dots, q_k$ to the path $p$, and let $c_0=r_p$.  
 %Let $d_i = d(c_i, c_{i+1})$, and let $d_k$ be the distance from $c_k$ to the leaf of $p$.  
 %Observe that $\sum_{i=0}^k d_i = l_p$.
 %
 % 
 %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
 First observe that for all $i$, $r_{q_i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_0$, as $L(T_{r_{q_i}})\subseteq L(T_{r_p})$.

 Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{q_i}$ and $r_{q_j}$ are not on the same root to leaf path.
 In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=1}^k |H_i\setminus H_i'| \leq |p|$ 
 as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
 Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_0$ we have $\sum_{i=1}^k |H_i'|\leq |H_0|$.
 Putting these together gives
 \[
  \sum_{i=1}^k |H_i| = \sum_{i=1}^k |H_i'| + \sum_{i=1}^k |H_i\setminus H_i'| \leq |H_0|+|p|.
 \]
\end{proof}

\begin{corollary}
 Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
 Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
\end{corollary}
\begin{proof}
 Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
 $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
 Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_0|+|p| = O(|p|)$, as $p$ is a heavy path. 
 Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.

@@

For a given coloring $\chi$ of a merge tree $T$, we now describe a particular path decomposition such that 
$f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.  By \Lem{algCost} and \Lem{pathDecomp} this will imply the running time of our algorithm 
is optimal with respect to the shape of the merge tree.

%We define two cost functions over validly colored trees.  To do so, we first define two quantities for each vertex $v\in T$.  
%First let $P_v$ be the longest path from $v$ to a descendant leaf of $v$.  Second, let $H_v$ be the subset of 
%vertices along the path from the root to $v$ such that one of the two colors  is in $\chi(L(T_v))$.

>>>>

Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
of this section proving that this running time is optimal.

@@

\begin{lemma}
\label{lem:pathBounds}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Then, 
 \begin{compactenum}
  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
 \end{compactenum}
\end{lemma}
\begin{proof}
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
 As such, 
 \[
 \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
 \]
 If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
 $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
\end{proof}

@@

 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
 Then $\sum_{i=1}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
 %Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
\end{lemma}
\begin{proof}
 %Let $c_1, \dots, c_k$ be the connections of the paths $q_1, \dots, q_k$ to the path $p$, and let $c_0=r_p$.  
 %Let $d_i = d(c_i, c_{i+1})$, and let $d_k$ be the distance from $c_k$ to the leaf of $p$.  
 %Observe that $\sum_{i=0}^k d_i = l_p$.
 %
 % 
 %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

 Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
 In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=1}^k |H_i\setminus H_i'| \leq |p|$ 
 as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
 Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=1}^k |H_i'|\leq |H_p|$.
 Putting these together gives
 \[
  \sum_{i=1}^k |H_i| = \sum_{i=1}^k |H_i'| + \sum_{i=1}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
 \]
\end{proof}

\begin{corollary}
 Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
 Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
\end{corollary}
\begin{proof}
 Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
 $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
 Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
 Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.

@@

For a given valid coloring $\chi$ of a tree $T$, we now describe a particular path decomposition for which 
we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about decompositions from the 
previous subsection.  By the above lemma this will imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.

\begin{definition}
 Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
 $T$ as follows.  For each internal vertex $v\in T$ add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
 where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
 The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
 we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
\end{definition}

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$.  Call a light path in $P(T)$ reachable from $p$ if there is a path in the tree
 $T$ from some vertex in $p$ which only moves to vertices of larger height and only vertices which lie in light paths.
 Call this set of reachable paths the \emph{light path tree} of $p$.  Naturally this set can be viewed as a tree 
 (i.e. $p$ is the root of this tree and each other path has a well defined depth from $p$).
\end{definition}

\begin{lemma}
 Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
 Let $p$ be a light path in $\mathcal{P}(T)$ and let $LPT(p)$ be the light path tree of $p$.

 Then $\sum_{q\in L(p)} \sum_{v\in q} \log |H_v| = |H_p|\log |H_p|$.
\end{lemma}
\begin{proof}UNDER CONSTRUCTION / NOTES:

 Claim: Let $q$ be some path of depth $i$ in $LPT(p)$, and let $H_q$ be the heap at the root of this path.
 Then $|H_q|< |H_p|/c^i$ for some constant $c>1$.  This should follow from using a maximum path decomposition.

 This claim would imply the depth of the tree is $O(\log |H_p|)$ and the total number of paths in the tree is $O(|H_p|)$, 
 thought I am not sure if either of these statements are useful.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Junk, don't read}
The following can be viewed as stronger version of \Lem{adjacent}, which is possible as the path is now assumed to be maximum.

\begin{lemma}
 Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
 Let $p$ be a light path in $\mathcal{P}(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.

 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
  Then $\sum_{i=1}^k |H_i| \leq |H_p|/c$ for some constant $c>1$, where $H_p = H_{r_p}$.
\end{lemma}

% \begin{proof}UNDER CONSTRUCTION: 
% For simplicity, let $H_{k+1}$ denote the heap at the leaf vertex of $p$.  As the $H_k+1$ is disjoing from all other $H_{i}$, \Lem{adjacent} implies that 
% $H_p > \sum_{j=1}^{k+1} H_i - |p|$.  
% 
% Let $x$ be the largest index such that $H_x > H_p / 3$.
% By the above observation and since $p$ is a light path we have 
% \[
%  H_p > \sum_{j=1}^{k} H_i + H_{k+1} - |p|
% \] 
% \end{proof}
